---
title: "In-class template"
author: "Lukas Lescano"
format: pdf
---

# 1. Set up

## Packages

```{r}
#| label: packages

# general use and cleaning
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# reading .xlsx files
library(readxl)

# visualize missing data
library(naniar)

# arrange plots
library(patchwork)
```


## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects")

# site information
sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites")

# water quality
water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality",
  na = c("", "n/a", "N/A", "over", "173+"))
```

# 2. Cleaning

## `sites` object

Goal: to create a new object from `sites` that only includes NCOS sites

```{r}
#| label: NCOS sites

# creating new object from sites
ncos_sites <- sites |> 
  # filtering to only include NCOS quarterly sampling sites
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")

```

## clean `taxon_list`

Goal: convert character strings in the existing `taxon_list` data frame into snakecase
(clean_names() turns all into snakecase)

Use mutate to manilulate anything in a column

```{r}
#| label: cleaning taxon

# creating new object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converted taxon names to snake case
  mutate(verbatim_name = to_any_case(verbatim_name,
                                     case = "snake"))
```

## clean `water_quality`

Goal: create a data frame with median pH, median DO (mg/L), and median salinity (ppt) for each data and site

- create a new object from water_quality
- clean column names using janitor
- only include surveys from quaterly ncos sites
- create a new column called `date_site` to join with later data frames
- group by date_site
- calculate median pH, salinity, and dissolved oxygen
- ungroup data frame
- filter out any outliers

semi_join (join used to filter)
since using pipe, dont need to provide left join data frame
 
```{r}
#| label: cleaning water_quality

# creating a new object from water_quality
water_quality_clean <- water_quality |> 
  # cleaning column names
  clean_names() |> 
  # used filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |>   
  # creating new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by sites
  group_by(date_site) |>
  # calculating median pH, salinity, and dissolved oxygen
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungrouping data frame
  ungroup() |> 
  # filtering out salinity outliers
  filter(date_site != "2022-08-12_NVBR")
```


## clean `aq_ins`

Goal: create a long format data frame in which each row contains the counted individuals in each taxon during each survey. The data frame should include taxonomic information for each organism and the water quality information for each survey.

RStudio functions expect long data frame not wide data frame

- create new object from `aq_ins`
- clean column names
- filter to only include quarterly NCOS sites
- filter to only include "filtered beaker" samples
- convert the data frame to long format
- group by site, date, and taxon
- calculate average abundance per liter (sum counts divided by 7.5 liters)
- create a new column called `date_site` to join with water quality
- join data frames
- join with taxon list

rename(new name = old name)

```{r}
#| label: cleaning aquatic insects

aq_ins_clean <- aq_ins |> 
  clean_names() |> 
  semi_join(ncos_sites, by = "site") |> 
  rename(date = date_on_vial) |> 
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  filter(sample_type %in% c("FB250", "FB 250")) |> 
  pivot_longer(
    # columns to make longer
    cols = ostracod:annelida_polychaete,
    # new column for the old column names
    names_to = "taxon",
    # new column for the values in each of the old columns
    values_to = "abundance"
  ) |> 
  group_by(date, site, taxon) |> 
  summarize(ave_lit = sum(abundance, na.rm = TRUE) / 7.5) |> 
  unite("date_site", date, site, remove = FALSE) |> 
  left_join(water_quality_clean, by = "date_site") |> 
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
  ))
```

